---
title: "<b>Predicting the 2026 Brownlow Medal</b>"
author: "Cooper Denny"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# **Introduction**

In this document, I constructed an ordinal logistic regression model to predict the 2026 AFL Brownlow Medal votes. To do so, I used a dataset I compiled from various trusted sources, including the fitzRoy R package, afl.com.au, AFL Tables, and the AFL Coaches Association. I took the time to prepare and clean the data, transforming it to lay a solid foundation for building a robust predictive model.

This is now the fourth year I've run this model, and the underlying approach has stayed largely consistent since its first outing in 2023. The one notable change came in 2024, when I added a `Position` variable. This states a player's primary position according to the AFL website, and I added it in after finding that some non-midfielders - particularly ruckmen such as Rowan Marshall, Max Gawn and Tim English - had very inflated predicted votes compared to what they actually polled in 2023. As the Brownlow Medal is considered a midfielder award (for good reason), I needed to incorporate playing position in some capacity. That same approach carried through into the 2025 model and again into this 2026 version.

For my 2026 Brownlow prediction modelling, I trained the model on home-and-away season data from 2015-2025, with the goal to forecast how many Brownlow votes a player might receive based on their performance metrics.

Once the model was trained, I shifted to the testing phase, where I applied it to player data from the 2026 season to generate game-by-game vote predictions. Unlike my 2023, 2024 and 2025 write-ups, this document doesn't compare those predictions against the official Brownlow votes - the 2026 count hasn't happened yet, as it's held after the home-and-away season concludes. Instead, this is purely a look at who the model expects to poll well, which I'll revisit once the umpires' votes are officially announced.

Let’s dive in!

# **Data Preparation**

## <u>Libraries</u>

To start, I loaded the key libraries needed for data analysis and modelling. These libraries were essential for handling tasks like data manipulation, statistical modelling, and formatting the output:

```{r}
library(tidyverse)   #Collection of R packages for data science
library(ordinal)     #Package for fitting ordinal regression models
library(knitr)       #Package for dynamic report generation in R
library(unglue)      #Package for string manipulation
library(fastDummies) #Package for creating dummy variables
library(writexl)     #Package for exporting data to Excel files
library(DT)          #Package for creating interactive tables
library(caret)       #Package to help create a confusion matrix 
library(ggplot2)     #Package for visual analysis and plotting
```

## <u>Data Pre-Processing</u>

I began by reading in the AFL Brownlow Medal data. The dataset includes detailed player statistics and was sourced from reliable outlets like the `FitzRoy` R package, afl.com.au, AFL Tables, and the AFL Coaches Association (AFLCA). I have already gathered, joined and cleaned the data into the file `brownlow_data_2026.csv` so that it is ready for analysis!

```{r}
#Read in the data
brownlow_data <- read.csv("brownlow_data_2026.csv")
```

To get a better understanding of the dataset, I took a quick look at the first 100 rows:

```{r}
#Print the table in the nice format
datatable(head(brownlow_data,100), options = list(scrollX = TRUE, pageLength = 5))
```

I also displayed the column names to get a sense of the variables included in the dataset:

```{r}
#Display the column names
colnames(brownlow_data)
```
There are `r length(colnames(brownlow_data))` variables that capture different aspects of each player's game performance. To simplify the analysis and avoid redundancy, I created new variables and remove overlapping ones. For instance:

- The `Disposals` variable isn’t needed since we already have `Kicks` and `Handballs` separately.
- I’ll create an `Ineffective.Kicks` variable by subtracting `Effective.Kicks` from `Kicks`, and then remove the original `Kicks` variable.

Now, let’s go ahead with these transformations and other necessary pre-processing steps:

```{r}
#Add a score involvements variable that doesn't include direct goals, behinds or goal assists
brownlow_data$Involvements.No.Scores.Or.Assists <- brownlow_data$Score.Involvements - 
                                                brownlow_data$Goals - 
                                                brownlow_data$Behinds - 
                                                brownlow_data$Goal.Assists

#Add a hitouts variable that doesn't include those that go to advantage
brownlow_data$Hitouts.No.Advantage <- brownlow_data$Hitouts - 
                                      brownlow_data$Hitouts.To.Advantage

#Add an Outside F50 ground ball gets variable 
brownlow_data$Outside.50.Ground.Ball.Gets <- brownlow_data$Ground.Ball.Gets - 
                                             brownlow_data$F50.Ground.Ball.Gets

#Add a contested offence 1-on-1 variable that doesn't include those won
brownlow_data$Contest.Offence.One.On.Ones.Not.Won <- brownlow_data$Contest.Offence.One.On.Ones - 
                                                     brownlow_data$Contest.Offence.Wins

#Add a contested defence 1-on-1 variable that doesn't include those lost
brownlow_data$Contest.Defense.One.On.Ones.Not.Lost <- brownlow_data$Contest.Defense.One.On.Ones - 
                                                      brownlow_data$Contest.Defense.Losses

#Add an ineffective kicks variable
brownlow_data$Ineffective.Kicks <- brownlow_data$Kicks - 
                                   brownlow_data$Effective.Kicks

#Add an ineffective handballs variable
brownlow_data$Ineffective.Handballs <- brownlow_data$Handballs - 
                                       brownlow_data$Effective.Handballs

#Add a shots at goal variable that doesn't include those that go in for a goal or behind
brownlow_data$Shots.No.Score <- brownlow_data$Shots.At.Goal - 
                                brownlow_data$Goals - 
                                brownlow_data$Behinds

#Add a forward half pressure acts variable
brownlow_data$Forward.Half.Pressure.Acts <- brownlow_data$Pressure.Acts - 
                                            brownlow_data$Defensive.Half.Pressure.Acts

#Add a kick ins variable that doesn't include those where the kicker plays on out of the goal square
brownlow_data$Kick.Ins.Not.Play.On <- brownlow_data$Kick.Ins - 
                                      brownlow_data$Kick.Ins.Play.On

#Add a contested possessions variable that doesn't include contested marks
brownlow_data$Contested.Possessions.No.Mark <- brownlow_data$Contested.Possessions - 
                                              brownlow_data$Contested.Marks

#Add an uncontested possessions variable that doesn't include uncontested marks
brownlow_data$Uncontested.Possessions.No.Mark <- brownlow_data$Uncontested.Possessions - 
                                                brownlow_data$Uncontested.Marks

#Add an intercepts variable that doesn't include intercept marks
brownlow_data$Intercepts.No.Mark <- brownlow_data$Intercepts - 
                                   brownlow_data$Intercept.Marks

#Add an intercepts variable that doesn't include intercept marks
brownlow_data$Tackles.Outside.50 <- brownlow_data$Tackles - 
                                    brownlow_data$Tackles.Inside.50

#Add a Player variable which pastes a player's first name and surname
brownlow_data$Player <- paste(brownlow_data$First.Name, brownlow_data$Surname)
```

To boost the predictive power of the model, I also added variables that capture votes from previous seasons. Players who have polled well in the past often continue to perform strongly.

```{r}
#Previous season Brownlow Votes and Coaches Votes variables
prev_season_votes <- brownlow_data %>% 
  group_by(Player.ID, Year) %>% 
  summarise(
  Brownlow.Votes.Previous.Season = sum(Brownlow.Votes, na.rm = TRUE),
  Coaches.Votes.Previous.Season = sum(Coaches.Votes, na.rm = TRUE)) %>% 
  mutate(Year = Year + 1)

#Join these new variables to the brownlow_data
brownlow_data <- full_join(brownlow_data, 
                           prev_season_votes) %>% 
                 filter(!is.na(First.Name))

#Replace NA Brownlow Votes with a 0
brownlow_data$Brownlow.Votes.Previous.Season <- replace(brownlow_data$Brownlow.Votes.Previous.Season, 
                                           is.na(brownlow_data$Brownlow.Votes.Previous.Season), 
                                           0)

#Replace NA Coaches Votes with a 0
brownlow_data$Coaches.Votes.Previous.Season <- replace(brownlow_data$Coaches.Votes.Previous.Season, 
                                                   is.na(brownlow_data$Coaches.Votes.Previous.Season), 
                                                   0)
```

I’ve also calculated the maximum votes a player received in any season prior to the one in which the game was played (dating back to 2015)

```{r}
#Max Brownlow Votes and Coaches Votes in a season variables

#Add empty variable 
Max.Brownlow.Votes.Season_prior <- data.frame()

#Loop through each season of available data to get the most Brownlow 
#Votes and Coaches Votes a player has received prior to that year
for(i in min(brownlow_data$Year):max(brownlow_data$Year)){
Max.Brownlow.Votes.Seasons_prior_i <- brownlow_data %>% 
  filter(Year <= i) %>%
  group_by(Player.ID, Year) %>% 
  summarise(Brownlow.Votes.Previous.Season = sum(Brownlow.Votes, na.rm = TRUE),
            Coaches.Votes.Previous.Season = sum(Coaches.Votes, na.rm = TRUE)) %>% 
  group_by(Player.ID) %>%
  summarise(Max.Brownlow.Votes.Season = max(Brownlow.Votes.Previous.Season),
            Max.Coaches.Votes.Season = max(Coaches.Votes.Previous.Season),
            Year = i + 1)

Max.Brownlow.Votes.Season_prior <- bind_rows(Max.Brownlow.Votes.Season_prior, Max.Brownlow.Votes.Seasons_prior_i)
}

#Join these new variables to the brownlow_data
brownlow_data <- full_join(brownlow_data, Max.Brownlow.Votes.Season_prior) %>% 
                 filter(!is.na(First.Name))

#Replace NA Brownlow Votes with a 0
brownlow_data$Max.Brownlow.Votes.Season <- replace(brownlow_data$Max.Brownlow.Votes.Season, 
                                          is.na(brownlow_data$Max.Brownlow.Votes.Season), 
                                          0)

#Replace NA Coaches Votes with a 0
brownlow_data$Max.Coaches.Votes.Season <- replace(brownlow_data$Max.Coaches.Votes.Season, 
                                                  is.na(brownlow_data$Max.Coaches.Votes.Season), 
                                                  0)
```

It was important to also convert the `Brownlow.Votes` and `Coaches.Votes` variables into factors. This ensures that R treats them as categorical rather than continuous variables, which is crucial for accurate interpretation and analysis when building the predictive model.

```{r}
#Convert Brownlow Votes variable to factor
brownlow_data$Brownlow.Votes <- as.factor(brownlow_data$Brownlow.Votes)

#Convert Coaches Votes variable to factor
brownlow_data$Coaches.Votes <- as.factor(brownlow_data$Coaches.Votes)
```

Next, I refined the dataset by keeping only the columns that are essential for the analysis.

```{r}
#Select variables that will be used in the model
brownlow_data <- brownlow_data %>% 
  select(Year, Game.ID, Round.Number, Team.Name, 
         Opponent.Name, Player, Goals, Behinds, 
         Effective.Kicks, Effective.Handballs, Ineffective.Kicks, Ineffective.Handballs, 
         Bounces, Tackles.Inside.50, Tackles.Outside.50, Contested.Possessions.No.Mark, 
         Uncontested.Possessions.No.Mark, Inside.50s, Marks.Inside.50, Contested.Marks, 
         Uncontested.Marks, One.Percenters, Clangers, Frees.For, 
         Frees.Against, Rebound.50s, Goal.Assists, Rating.Points, 
         Turnovers, Intercepts.No.Mark, Involvements.No.Scores.Or.Assists, Metres.Gained, 
         Centre.Clearances, Stoppage.Clearances, Marks.On.Lead, Intercept.Marks, 
         Hitouts.To.Advantage, Hitouts.No.Advantage, Outside.50.Ground.Ball.Gets, F50.Ground.Ball.Gets, 
         Score.Launches, Forward.Half.Pressure.Acts, Defensive.Half.Pressure.Acts, Spoils, 
         Contest.Offence.One.On.Ones.Not.Won, Contest.Offence.Wins, Contest.Defense.One.On.Ones.Not.Lost, 
         Contest.Defense.Losses, Shots.No.Score, Time.On.Ground.Percentage, Team.Goals, Team.Behinds, 
         Opponent.Goals, Opponent.Behinds, Margin, Kick.Ins.Not.Play.On, 
         Kick.Ins.Play.On, Max.Brownlow.Votes.Season, Max.Coaches.Votes.Season, Brownlow.Votes.Previous.Season, 
         Coaches.Votes.Previous.Season, Coaches.Votes, Team.Result, Position, Brownlow.Votes)
```

This updated dataset, with the newly created and refined variables, will serve as the foundation for developing the predictive model.

```{r}
#Present the enhanced dataset in a nice format
datatable(head(brownlow_data,100), options = list(scrollX = TRUE, pageLength = 5))
```

# **Model Preparation**

## <u>Training Data Selection</u>

To begin, I specified the time periods for training and testing the model. The training period covers data from 2015 to 2025, while the testing period focuses on the 2026 season:

```{r}
#First season of training period
train_start_season <- 2015

#Last season of training period
train_end_season <- 2025

#Season to predict Brownlow Votes
test_season <- 2026
```

Next, I created a training dataset by filtering the data for the specified seasons from 2015 to 2023.

```{r}
#Create dataset that encompasses the training period
brownlow_train <- brownlow_data %>% 
  filter(Year >= train_start_season, Year <= train_end_season)
```

## <u>Standardising the Numeric Statistics</u>

Next, I standardised the numeric statistics in the training dataset to ensure that each variable has a mean of 0 and a standard deviation of 1. This normalization helps improve the model's performance by ensuring that all variables are on a similar scale.

```{r}
#Standardise the numeric values
train_numeric_standardised <- scale(brownlow_train[,8:(ncol(brownlow_train)-4)])
brownlow_train[,8:(ncol(brownlow_train)-5)] <- train_numeric_standardised

#Store the center and scale as this will be used to standardise the testing set
train_numeric_standardised.center<-attr(train_numeric_standardised,"scaled:center")
train_numeric_standardised.scale<-attr(train_numeric_standardised,"scaled:scale")
```

## <u>Creating Dummy Variables</u>

Next, I converted categorical variables like `Coaches.Votes`, `Team.Result` and `Position` into dummy variables. These dummy variables are binary (0 or 1) and allow the model to process categorical data more effectively:

```{r}
#Create dummy variables for Coaches Votes and Team Result
brownlow_train <- dummy_cols(brownlow_train, 
                             select_columns = c('Coaches.Votes', 'Team.Result', 'Position'),
                             remove_selected_columns = TRUE)
```

# **Model Training**
## <u>Building the Ordinal Logistic Regression Model</u>

To predict the number of Brownlow votes a player might receive based on their performance statistics, I used an ordinal logistic regression model. This model takes into account various metrics from the cleaned and enhanced dataset.

### <u>Why an ordinal model?</u>

`Brownlow.Votes` only ever takes one of four values - 0, 1, 2 or 3 - and, critically, those values are ordered: 3 votes represents a better game than 2, which is better than 1, which is better than 0. This rules out two more obvious modelling choices:

- **Linear regression** would treat votes as a continuous number and assume the "distance" between 0 and 1 votes is identical to the distance between 2 and 3 votes. It would also happily predict nonsensical values like 1.7 or -0.4 votes.
- **Multinomial logistic regression** would treat 0, 1, 2 and 3 as unrelated, unordered categories - the same as if they were team names. It would throw away the information that 3 is "more" than 2, and it needs a lot more parameters to do so (a separate set of coefficients per category, rather than one shared set).

An **ordinal logistic regression** (fitted here as a cumulative link model, via `clm`) is built specifically for outcomes like this. Instead of modelling the probability of each vote count directly, it models the *cumulative* probability of scoring at most a given number of votes, using the same underlying logic as standard logistic regression but applied at every cut-point in the scale at once.

Concretely, for an outcome $Y$ with ordered categories $j = 0, 1, 2, 3$ and a vector of predictors $x$ (goals, tackles, coaches votes, etc.), the model assumes there's an unobserved, continuous "vote-worthiness" score for each player-game, and that a player crosses into a higher vote category once that score passes a threshold. This is expressed as a set of cumulative logits:

$$
\text{logit}\big(P(Y \le j \mid x)\big) = \log\left(\frac{P(Y \le j \mid x)}{1 - P(Y \le j \mid x)}\right) = \theta_j - \beta^\top x, \quad j = 0, 1, 2
$$

Here, each $\theta_j$ is a **threshold** (cut-point) on the latent vote-worthiness scale separating category $j$ from $j+1$ - these are the `Threshold coefficients` (`0|1`, `1|2`, `2|3`) printed in the model summary below. The $\beta$ coefficients are shared across all three cut-points; this is the **proportional odds assumption**, and it's what lets the model estimate just one coefficient per predictor (e.g. one coefficient for `Goals`) rather than a separate coefficient for every vote threshold. A positive $\beta$ for a predictor means that as it increases, a player is pushed towards the higher end of the vote scale at every cut-point simultaneously - which is exactly the kind of "more of this stat generally means more votes" relationship I'd expect for something like `Goals` or `Effective.Kicks`.

This structure fits the Brownlow voting process well: umpires aren't picking from four unrelated buckets, they're ranking the best performers in a game and then discretising that ranking into a 3-2-1 (plus 0 for everyone else) scale. The ordinal model's latent-threshold framing mirrors that process much more naturally than treating vote counts as either a plain number or an unordered label, while also letting me draw on the coefficients' statistical significance (via the `Coefficients` table's p-values) to see which performance metrics most reliably move a player up that latent scale.

The code below defines the model formula and fits the ordinal logistic regression model using the `clm` function from the `ordinal` package. This package is tailored for fitting cumulative link models, including ordinal logistic regression. The `clm` function estimates the model parameters through maximum likelihood estimation, which is well-suited for the ordered nature of the response variable. This is particularly important for ordinal outcomes, like the number of Brownlow votes, where the categories are ranked.


```{r}
my_model <- clm(Brownlow.Votes ~ 
                      Goals + Behinds + Effective.Kicks + Effective.Handballs + 
                      Ineffective.Kicks + Ineffective.Handballs + Bounces + 
                      Tackles.Inside.50 + Tackles.Outside.50 + Contested.Possessions.No.Mark + 
                      Uncontested.Possessions.No.Mark + Inside.50s + Marks.Inside.50 + 
                      Contested.Marks + Uncontested.Marks + One.Percenters + 
                      Clangers + Frees.For + Frees.Against + 
                      Rebound.50s + Goal.Assists + Rating.Points + 
                      Turnovers + Intercepts.No.Mark + Involvements.No.Scores.Or.Assists + 
                      Metres.Gained + Centre.Clearances + Stoppage.Clearances + 
                      Marks.On.Lead + Intercept.Marks + Outside.50.Ground.Ball.Gets + 
                      F50.Ground.Ball.Gets + Score.Launches + Forward.Half.Pressure.Acts + 
                      Defensive.Half.Pressure.Acts + Spoils + Contest.Offence.One.On.Ones.Not.Won + 
                      Contest.Offence.Wins + Contest.Defense.One.On.Ones.Not.Lost + Contest.Defense.Losses + 
                      Shots.No.Score + Time.On.Ground.Percentage + Team.Result_W + 
                      Team.Result_L + Team.Goals + Team.Behinds + 
                      Opponent.Goals + Opponent.Behinds + Kick.Ins.Not.Play.On + 
                      Kick.Ins.Play.On + Hitouts.To.Advantage + Hitouts.No.Advantage + 
                      Position_KEY_DEFENDER + Position_KEY_FORWARD + Position_MEDIUM_DEFENDER + Position_RUCK +
                      Position_MEDIUM_FORWARD + Position_MIDFIELDER + Position_MIDFIELDER_FORWARD + 
                      Coaches.Votes_1 + Coaches.Votes_2 + Coaches.Votes_3 + 
                      Coaches.Votes_4 + Coaches.Votes_5 + Coaches.Votes_6 + 
                      Coaches.Votes_7 + Coaches.Votes_8 + Coaches.Votes_9 + 
                      Coaches.Votes_10 + Max.Brownlow.Votes.Season + Brownlow.Votes.Previous.Season + 
                      Max.Coaches.Votes.Season + Coaches.Votes.Previous.Season,
                 
                 data = brownlow_train, link = "logit", #Use logistic link function
                 
                 threshold = "flexible", #Allow flexible thresholds
                 
                 Hess = TRUE, #Compute Hessian matrix for testing
                 
                 maxIter = 100, #Allow up to 100 iterations
                 
                 tol = 1e-6, #Set a small tolerance for convergence
                 
                 contrasts = list(Team.Name = "contr.treatment"), #Specify contrasts for the 'Team' variable
                 
                 na.action = na.omit)
```

## <u>Model Summary</u>

Finally, I summarised the model to evaluate its performance and understand the significance of each predictor variable.

This summary includes details about the coefficients, standard errors, z-values, and p-values for each predictor in the model. It helps to assess the significance of each predictor variable and gauge the overall performance of the model.

```{r}
#Print a summary of the model
summary(my_model)
```
The summary of the fitted ordinal logistic regression model for predicting Brownlow Votes provides a detailed look at the various performance metrics and their importance in determining vote counts.

Key offensive metrics, such as `Goals` (0.757), `Effective.Kicks` (0.517), and `Effective.Handballs` (0.492), show strong positive coefficients and highly significant p-values (all p < 2e-16). Of the standardised on-field performance metrics, `Goals` has the largest effect - each one standard deviation increase in a player's `Goals` for the game pushes them furthest up the latent vote-worthiness scale. This indicates that these actions greatly boost the chances of receiving votes and serve as reliable predictors. Interestingly, `Ineffective.Kicks` (0.343, p < 2e-16) and `Ineffective.Handballs` (0.182, p = 3.49e-14) are also positive and highly significant, just with smaller coefficients than their effective counterparts - suggesting umpires still credit general disposal involvement even when the disposal itself doesn't find a target.

Defensive metrics are more of a mixed bag than I expected. `Tackles.Outside.50` (0.158, p = 6.53e-16) is a strong, significant positive predictor, but `Tackles.Inside.50` (0.032, p = 0.072) only just misses conventional significance, suggesting forward-half tackling pressure isn't as reliably rewarded as tackling elsewhere on the ground. Similarly, `Intercept.Marks` (0.132, p = 1.10e-08) is a solid positive predictor, but `Intercepts.No.Mark` (intercept possessions that aren't marks) actually carries a small negative, significant coefficient (-0.063, p = 0.020) - it seems umpires reward the more eye-catching intercept mark far more than a routine intercept possession. `Frees.Against` (-0.025, p = 0.236) is not statistically significant in this model, so I can't say penalties measurably hurt a player's vote chances once everything else is accounted for.

Team performance metrics play a role, but not quite as symmetrically as I'd assumed. Losing a match (`Team.Result_L`: -0.744, p = 5.19e-05) has a clear, significant negative effect on vote-getting, but simply winning (`Team.Result_W`: 0.145, p = 0.427) is not itself significant - it's the losses that matter more than the wins. Both a team's own scoring (`Team.Goals`: -0.172, p = 2.91e-13) and its opponent's scoring (`Opponent.Goals`: -0.329, p < 2e-16) carry significant negative coefficients, which reflects that high-scoring shootouts tend to spread the votes more thinly across more contributors than tighter, lower-scoring contests.

As I assumed, the model also takes into account a player's primary position. `Position_MIDFIELDER` (0.339, p = 1.91e-11) is a strong, significant positive predictor relative to the reference position, while `Position_RUCK` (-0.306, p = 0.021) is significantly negative - confirming that midfielders remain more likely to receive votes than rucks. `Position_MEDIUM_DEFENDER` (-0.108, p = 0.198) trends negative but isn't statistically significant this time around, so I can't say with confidence that medium defenders are disadvantaged once the rest of the model's variables are held constant.

But most eye-opening, is that coaches votes have a significant relationship to vote-getting on Brownlow Medal night, with every level of coaches votes (from 1 to 10) coming back highly significant (all p < 2e-16) and the coefficients climbing steadily from 0.823 at `Coaches.Votes_1` up to 3.577 at `Coaches.Votes_10` - by far the largest coefficients in the entire model, dwarfing even `Goals`. It's worth noting these aren't quite comparable in a one-to-one sense: the on-field statistics are standardised, so their coefficients represent the effect of a one standard deviation change, whereas each `Coaches.Votes` coefficient represents the full jump from receiving zero coaches votes to receiving that many. Even accounting for that difference in scale, though, it's a clear, monotonic relationship - the more strongly the coaches rated a player's game, the more likely the umpires were to vote for them too, and coaches recognition looks to be the single most powerful signal in the model.

I've decided not to regularize or reduce the model because my primary goal was to capture as much detailed information as possible from the available data. Since the Brownlow Medal voting process is subjective and influenced by a multitude of factors, I wanted to ensure that no potentially valuable information was excluded from the model. Regularizing or simplifying the model could have risked losing some of this nuance, potentially missing out on subtle patterns that contribute to how votes are awarded. By allowing the model to be more complex, I aimed to increase its ability to pick up on these smaller details, which may be critical in accurately predicting votes in tight races or less obvious matchups. In this case, maximizing the model’s flexibility was more important than trying to prevent overfitting, as I believe the richness of the data outweighs the potential risks of over-complication.

Overall, this model emphasizes the complex nature of player performance evaluation, weaving together offensive skills, defensive capabilities, team success, and external recognition to predict Brownlow Votes. The combination of significant variables and their highly significant p-values suggests that excelling in multiple areas of the game is essential for gaining votes.

# **Model Testing**

The model testing process involves applying the trained model to a new season's data to predict the Brownlow Medal votes for each player in each game. This testing phase assesses the model's performance and ensures its predictions are reliable and accurate.

```{r}
#Create dataset that encompasses the testing period
brownlow_test <- brownlow_data %>% 
  filter(Year == test_season)

#Print the table in the nice format
datatable(head(brownlow_test,100), options = list(scrollX = TRUE, pageLength = 5))
```

After a quick peek of the 2026 data, I then standardise the numeric values on the same scale as the training set, as well as create dummy variables for the categorical variables `Coaches.Votes`, `Team.Result` and `Position`.

```{r}
#Standardise the numeric values, based on the training set
brownlow_test[,8:(ncol(brownlow_test)-4)] <- scale(brownlow_test[,8:(ncol(brownlow_test)-4)],
                                                   center=train_numeric_standardised.center,
                                                   scale=train_numeric_standardised.scale) 

#Create dummy variables for Coaches Votes, Team Result and Position
brownlow_test <- dummy_cols(brownlow_test, 
                            select_columns = c('Coaches.Votes', 'Team.Result', 'Position'),
                            remove_selected_columns = TRUE)
```

## <u>Game-By-Game Votes</u>

Next, the trained model is used to make predictions on the test data. The probabilities of receiving different vote counts (0, 1, 2, or 3 votes) are calculated for each player in each game.

```{r}
#Use the predict function to make predictions on the test data
predictions <- predict(my_model,
                       newdata = brownlow_test %>% select(-Brownlow.Votes), 
                       type = 'prob')

#Transform the predictions into a dataframe that is readable
predictions_probability_matrix  <- data.frame(matrix(unlist(predictions), 
                                              nrow = nrow(brownlow_test)))

#Change the column names of the predictions_probability_matrix dataframe
colnames(predictions_probability_matrix) <- c("Votes.0", "Votes.1", "Votes.2", "Votes.3")

#Bind the columns of brownlow_test dataframe with predictions_probability_matrix 
brownlow_test_predictions  <- cbind.data.frame(brownlow_test, predictions_probability_matrix)
```

Expected votes for each player for each game are then calculated:

```{r}
#Calculate the "expected" votes a player should receive each game according to the model
brownlow_test_predictions$Expected.Votes <- 1*brownlow_test_predictions$Votes.1 + 
                                            2*brownlow_test_predictions$Votes.2 + 
                                            3*brownlow_test_predictions$Votes.3
```

Let's see which players had the highest "Expected Votes" games according to the model:

```{r}
#Ordering player's games according to their "expected" votes
expected_votes <- brownlow_test_predictions %>% 
  select(Game.ID,	Round.Number,	Player, Team.Name, Opponent.Name, Expected.Votes) %>%
  arrange(desc(Expected.Votes))

#Presenting this table in a nice format
datatable(expected_votes, options = list(scrollX = TRUE, pageLength = 10))
```

So, according to the model, `r expected_votes$Player[1]`'s Round `r expected_votes$Round.Number[1]` performance against `r expected_votes$Opponent.Name[1]` is considered the best game of the season so far according to `Expected.Votes`, followed by a number of notable individual performances from other stars of the competition.

However, with the way the Brownlow Medal works, we can only give out a single 3 vote, a single 2 vote, and a single 1 vote each game. But by using the `Expected.Votes`, we can assign the 3, 2 and 1 based on the top three players `Expected.Votes`.

Since the 2026 Brownlow Medal votes aren't awarded until after the home-and-away season, there's no `Brownlow.Votes` column to compare these against just yet - that comparison will have to wait until the count.

```{r}
#Assign the 3, 2 and 1 votes to the top 3 "expected" votes players of each game
round.by.round.votes <- brownlow_test_predictions %>%
  group_by(Game.ID) %>%
  top_n(3, Expected.Votes) %>%
  mutate(Predicted.Votes = order(order(Expected.Votes, Player, decreasing=FALSE))) %>%
  select(Game.ID, Round.Number, Player, Team.Name,
         Opponent.Name, Predicted.Votes, Expected.Votes) %>%
  arrange(Game.ID, desc(Predicted.Votes))
```

Let's display our predicted votes for each game in a nice format:

```{r}
#Filter round.by.round.votes to only see players predicted to poll a vote in each game
game_vote_predictions <- round.by.round.votes %>%
                         filter(Predicted.Votes %in% c(1,2,3))

#Present the table in a nice format
datatable(game_vote_predictions, options = list(scrollX = TRUE, pageLength = 3))
```

## <u>Season Votes</u>

I can then put the round-by-round votes altogether into a predicted leaderboard based off of the model votes.

```{r}
summary_table <- game_vote_predictions %>%
  ungroup() %>% select(Player, Team.Name, Round.Number, Predicted.Votes) %>%
  pivot_wider(names_from = Round.Number, values_from = Predicted.Votes, names_prefix = "R.") %>%
  mutate(across(starts_with("R."), as.numeric)) %>%
  mutate(
    Predicted_Votes = rowSums(select(., starts_with("R.")), na.rm = TRUE),
    Count_3_Votes = rowSums(select(., starts_with("R.")) == 3, na.rm = TRUE),
    Count_2_Votes = rowSums(select(., starts_with("R.")) == 2, na.rm = TRUE),
    Count_1_Vote = rowSums(select(., starts_with("R.")) == 1, na.rm = TRUE),
    Games_Polled = rowSums(select(., starts_with("R.")) > 0, na.rm = TRUE)
  ) %>% 
  ##select(1:2, 28, 3:27, 29:32) %>%
  arrange(desc(Predicted_Votes), desc(Count_3_Votes), desc(Count_2_Votes), desc(Count_1_Vote))

datatable(summary_table, options = list(scrollX = TRUE, pageLength = 10))
```

This is my projected Brownlow Medal leaderboard for 2026, built purely from the model's game-by-game predicted votes. It's worth noting the leader here, `r summary_table$Player[1]`, currently sits atop the table with `r summary_table$Predicted_Votes[1]` predicted votes - but as always, a handful of close-fought games either way can shuffle the top of the count.

```{r}
#Create a vote tally based on the predicted votes of the model
vote_tally <- game_vote_predictions %>%
  group_by(Player, Team.Name) %>%
  summarise(Predicted.Votes = 3 * sum(Predicted.Votes == 3) +
                              2 * sum(Predicted.Votes == 2) +
                              1 * sum(Predicted.Votes == 1), .groups = 'drop') %>%
  arrange(desc(Predicted.Votes))

#Present the table in a nice format
datatable(vote_tally, options = list(scrollX = TRUE))
```

Since the 2026 Brownlow Medal count is still to come, there's no MAE, RMSE or predicted-vs-actual correlation to report this time around - that comparison, along with an updated results table, will follow once the umpires' votes are officially announced, just as it has for the 2023, 2024 and 2025 models.

# **Conclusion**

In conclusion, the development of this ordinal logistic regression model for predicting Brownlow Medal votes continues to show its usefulness in providing reliable, well-reasoned predictions based on a solid dataset of player performances. By incorporating a range of offensive and defensive statistics, the model captures key aspects of player performance, allowing for a nuanced understanding of what contributes to Brownlow recognition. Key metrics like `Goals`, `Effective.Kicks` and `Effective.Handballs` are the strongest predictors of votes, while defensive contributions such as `Tackles.Outside.50` and `Intercept.Marks` also play a significant, if more modest, role.

Furthermore, the model's integration of Coaches' votes further enhances its predictive power, with every level from 1 to 10 coming back highly significant - players recognised by the coaches are consistently more likely to receive umpire votes too. Past recognition is a more mixed story: a player's `Brownlow.Votes.Previous.Season` is a significant predictor, but a player's `Coaches.Votes.Previous.Season` and both `Max.Brownlow.Votes.Season` and `Max.Coaches.Votes.Season` are not significant in this fitted model, suggesting recent form carries more weight than a player's career-best season.

Across its previous three applications, the model has posted a Mean Absolute Error of 2.165 in 2023, 2.008 in 2024, and 2.228 in 2025, with 3-vote accuracy in the mid-to-high 50s each year. It correctly picked Patrick Cripps as the 2024 Medallist after predicting Lachie Neale to finish third (rather than win) in 2023, and in 2025 its top pick, Nick Daicos, ultimately finished runner-up to the actual winner, Matt Rowell, whom the model had ranked only sixth. This is now the fourth time the same approach has been applied, with the training data simply growing by another season each year.

Once the 2026 Brownlow Medal votes are counted, I'll come back and update this document with the game-by-game and season-wide comparisons against the official results, in keeping with prior years. Until then, this remains a forecast rather than a verdict - but it's one built on the same foundation that has, so far, proven a reasonably reliable guide to how the umpires see the game.
